Brownian motion of main-belt asteroids on human timescales

### 2025-02-22 main.tex
Yarkovsky text

One of our tasks will be to derive, from the known asteroid population, the rate (per target) of encounters vs the imparted impulse,
\begin{equation}
  \frac{dN}{dt\,dI_r\,dI_\phi\,dI_z},
  \label{eq:dN}
\end{equation}
where the components of $\vecI$ are given in the cylindrical basis
vectors about the asteroid's position at the impulse.
We will estimate this function from the known population,
approximating it as constant within each of three subsets of tracer MBAs:
the inner, middle, and outer regions bounded by the 4:1, 3:1,
5:2, and 2:1 mean-motion resonances with Jupiter.
We will further assume that the impulses on a given target are drawn
independently from this distribution, \ie\ a Poisson process defined
by this rate.  In this case, the only properties of the impulse
distribution that we need are its second moments for components $\alpha \in \{r,\phi,z\}$
\begin{equation}
  \left \langle n I_\alpha^2 \right\rangle \equiv \int d^3I \frac{dN}{dt\,dI_r\,dI_\phi\,dI_z} I_\alpha^2.
\label{eq:nvsq}
\end{equation}
The average of the cross terms $I_rI_z$ and $I_\phi I_z$ vanishes if
the deflector distribution is symmetric in inclination as expected, and we find numerically that the mean $I_rI_\phi$ is small enough to ignore.

From this knowledge of the impulse distribution, our goal is to obtain
the covariance matrix of the deviations of the target's position from
the initial orbit, after some time $T.$ The position observables are
the range, plus the celestial latitude and longitude of the target.  We will simplify our results by assuming a heliocentric observer, so that the observational position vector is $\vecp\equiv (r,\theta=z/r,\phi).$ The quantity we seek is the covariance matrix $\covm^p$ of the observations attributable to the accumulated gravitational perturbations:
\begin{equation}
  \covm^p  \equiv  \left\langle  \Delta\vecp \Delta\vecp^T\right\rangle,  \label{eq:Cp}
\end{equation}
where the angle brackets indicate an average over possible
realizations of the impulse history, and over the mean anomaly $M$ at the time of observation.  To do so, we will introduce an intermediate set of 6 parameters $\vecq$ describing the deviations of the orbital elements induced by the impulses.  The $\vecq$ components will be selected to be equal to zero before any impulses are applied, and each responds linearly to $\vecI$ at $|I|\ll 1.$
We will derive the matrix $\matA$ that describes the orbital-element shifts at time $T$ that arise from an impulse at time $t_i$: 
\begin{equation}
  \Delta \vecq(T,t_i) = \matA(e_0, T, t_i) \cdot \vecI.
  \label{eq:A}
\end{equation}
In our first-order perturbation theory, $\vecq(T)$ will be the sum of the $\Delta\vecq_i$ imparted by all impulses applied at times $0<t_i<T.$  Because the impulses are uncorrelated, $\vecq$ will therefore be the result of a random walk.  The distribution will have a covariance matrix $\covm^q(T) \equiv \left\langle \vecq(T) \vecq^T(T) \right\rangle$ whose elements are
\begin{eqnarray}
  \covm^q_{jk}(e_o,T) & = & \left\langle \sum_{i,\gamma} \matA_{j\gamma}(e_0,T,t_i) \matA_{k\gamma}(e_0,T,t_i) I^2_{i,\gamma} \right\rangle \\
           & = & \sum_\gamma \int dt_i \matA_{j\gamma}(e_0,T,t_i) \matA_{k\gamma}(e_0,T,t_i) \left\langle n I_\gamma^2\right\rangle.
\label{eq:Cqjk}
\end{eqnarray}
In the first line, the sum $i$ runs over the impulses and $\gamma$ runs over the components $r,\phi,z$ of the impulse. The second line evaluates the expectation value of averaging over realizations of the random walk of impulses, exploiting the independence of the individual impulses from each other.

The last element of our calculation will be a conversion from the
element shifts $\vecq$ into the observed displacements $\Delta\vecp$
at time $T.$  In linear perturbation theory this will again be expressible as a matrix
\begin{equation}
  \Delta\vecp(e_0,T) = \matB(e_0,T) \cdot \vecq(T).
\label{eq:B}
\end{equation}
Combining this with \eqq{eq:Cp} and \eqq{eq:Cqjk} yields the desired result
\begin{eqnarray}
  \covm^p_{\alpha\beta}(e_0,T) & = & \sum_{jk} B_{\alpha j}(e_0,T) B_{\beta k}(e_0,T) \covm^q_{jk}(T) \nonumber \\
  & = & \sum_\gamma \left\langle nI_\gamma^2\right\rangle \int dt_i \left[AB(e_0,T,t_i)\right]_{\alpha\gamma}  \left[AB(e_0,T,t_i)\right]_{\beta\gamma} 
\label{eq:ABq}
\end{eqnarray}
which we would wish to average over the phase of the initial mean anomaly $M_0$ of the MBA.

@@

\section{Impulse distribution}
\label{sec:impulse}
\subsection{Derivation}
Under the assumption that the distributions of masses and orbital
elements of MBAs are separable, the quantities needed to describe the
random walk of the asteroid's orbits can be expressed as an integral
over the absolute magnitude $H$ of the deflectors, the velocity $v$,
impact parameter $b,$ and polar/azimuthal angles $\theta,\phi$ of the
unit vector of the impulse direction $\bhat$:
\begin{eqnarray}
  \left\langle n I_\gamma^2 \right\rangle & = & \int
                                              dH\,dv\,db\,d\theta\,d\phi
                                             \frac{dN}{dH} \frac{dn}{dv\,db\,d\theta\,d\phi}
                                              I^2_\gamma(H,v,b,\theta\phi)
  \\
  & = & \int dH \frac{dN}{dH} \int dv
        \frac{dn}{dA\,dv\,d\theta\,d\phi} \hat b_\gamma^2(\theta,\phi)
        \int_{b_{\rm min}}^{b_{\rm max}} 2\pi b\,db\,
        \left(\frac{\sigma_M(H)}{bv}\right)^2 \\
  & = & 2\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int dv
        \frac{dn}{dA\,dv} v^{-2} \left\langle\hat b_\gamma^2(v)\right\rangle
        \log(b_{\rm max}/b_{\rm min}).
\end{eqnarray}
The second row adopts the (numerically verified) assumption that the
distribution of impact parameter $b$ will always be $\propto dA=2\pi
b\,db$ in the range of $b$ of interest to us.  The mean rate of encounters
between a deflector-tracer pair, as a function of the impact velocity
and geometry, is $dn/dA\,dv\,d\theta\,d\phi,$ which we will determine
through numerical measurements with the orbits of all known MBAs. The
$H$ distribution of MBAs is $dN/dH.$ We denote as $\sigma_M(H)$
the RMS error in the mass of MBAs at $H$ from the value used in the ephemeris model.
We use $\sigma_M$ rather than $M$ since we are seeking the size of
deviations from the ephemeris model.  For MBAs that are not included
in the ephemeris model, $\sigma_M=M(H),$ the full mass of the deflector.

@@

If we crudely choose some threshold
$I_{\rm det}$ of impulse as being large enough to generate high-$S/N$
deflections on its tracer, then our condition becomes $M(H)/bv <
I_{\rm det},$ or $b_{\rm min}=M(H)/vI_{\rm det}.$  We then have
\begin{equation}
   \left\langle n I_\gamma^2 \right\rangle = 
2\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int_{\sqrt{M(H)/I_{\rm
      det}}}^\infty \frac{dv}{v^2} 
        \frac{dn}{dA\,dv}  \left\langle\hat b_\gamma^2(v)\right\rangle
        \log \left[v^2 I_{\rm det}/M(H)\right].
        \label{eq:nI2}
      \end{equation}
The lower bound on the $v$ integral marks the speed below which
$b_{\rm max}<b_{\rm min}.$  Since $b_{\rm min}$ and $b_{\rm max}$
appear logarithmically, the estimated diffusion is relatively insensitive to choices of
their values.

@@

From this full MBA list, we select a random subset of 5000 objects
from each of the Inner, Middle, and Outer belts to serve as a sample
of ``tracers.''  We record all of the passages of any tracer MBA
within 0.03~AU of any other ``deflector'' MBA (drawn from the full
catalog of 1.26~million) over a 10-year period.  The 5000 tracers per
belt region are enough to be representative of the statistics of
the region, but few enough that the calculation can be done easily
on a laptop computer.
We will estimate the constituents of  $\langle nI_\gamma^2\rangle$ in
\eqq{eq:nI2} separately for each of the nine combinations of tracers
in Inner, Middle, and Outer regions with deflectors from each region.

@@

The leapfrog integrator advances all 1.26 million state vectors by 2
days, and the process is repeated.  We continue searching for pairs at
2-day intervals until we have recorded 10 years' worth of encounters---17.7 million
total impulses.

@@

For each pair of tracer-deflector regions (\eg\ Inner-Inner), we combine the list of
events with the counts of candidate tracers and deflectors to
calculate $dn/dA\,dv,$ the event rates per eligible pair; and the mean
geometry of the encounter, $\langle \hat b^2_\gamma \rangle,$ as a
function of $v$.  Figure~\ref{fig:rtz} plots these functions of $v$
for the Inner-Inner, Middle-Middle, and Outer-Outer scattering
events.  Here it is apparent that the $\hat b_\phi$ component is
larger than $\hat b_z$ and $\hat b_r,$ but there is significant
variation with MBA class and with $v$.  An important result is that
the quantity $v^{-2} (dN/dA\,dv)$ is always well-behaved as $v\rightarrow 0,$
so we do not need any special treatment to sum the integrals in
\eqq{eq:nI2}.  The $v$ integrand peak is in the range $0.03<v/v_c<0.1.$

@@

To give some context, a
comprehensive investigation of MBA masses from mutual scattering
events by \citet{goffin} reports typical $\sigma_M\approx
5\times10^{-13}$ for 30 MBAs. But we can expect substantial improvement
in the LSST era when a millions of potential tracer asteroids attain
mas-level astrometry. \citet{negin} use a simulated LSST sequence of observations to estimate the impulse (and mass) uncertainties from mutual encounters of the known asteroids that will occur during years 2--9 of its 10-year survey.  The tracer MBAs for those events are essentially an ubiased selection from the known MBAs, and the left-hand plot of Figure~\ref{fig:sigmaM} shows the distribution of the uncertainty $\sigma_I$ derived from each tracer's LSST data after marginalizing over its initial state vector.  This histogram marginalizes over the apparent magnitude and observing conditions of the tracer, the date of the encounter, and the direction $\bhat$ of the impulse.  The right-hand plot of Figure~\ref{fig:sigmaI} forecasts the distribution of $\sigma_M$ that we can expect for deflectors.  This forecast is made as follows: for each of the 5000 deflectors that we have selected from each MBA region, we have a list of all of its encounters with known MBAs with $b<0.03$~AU.  For each such encounter, we select at random one $\sigma_I$ from the distribution found by \citet{negin}, and calculate $\sigma_M=(bv/2)\sigma_I.$  For each deflector, we then plot the histogram of (a) the individual encounter with lowest $\sigma_M$; (b) the combined constraint $\left(\sum \sigma_{M,i}^2\right)^{-1/2}$ from the 10 lowest $\sigma_M$ encounters for that deflector; and (c) the combined constraint for all $b<0.03$~AU encounters (typically $\sim1000$).  The RMS $\sigma_M$ using LSST data from the 10 most-informative tracer encounters with a given deflector is estimated as $\sigma_M=2.6\times10^{-14}M_\odot.$ There are $\approx450$ MBAs with mass above this forecasted RMS $\sigma_M.$  This is a conservative estimate in that we ignore any observational information before LSST, and we also do not consider information from the $\approx10\times$ more MBAs that LSST will discover.

\subsubsection{Individually detectable impulses}
Examining the $\sigma_I$ distribution forecasted for LSST data on tracers (Figure~\ref{fig:sigmaI}, we find that for impulses occurring near the midpoint of the survey, $\approx80\%$ of all cases yield $\sigma_I<10^{-4}\,{\rm m}\,{\rm
  s}^{-1},$ some an order of magnitude lower.  We conservatively
assume that any encounter producing  an impulse $>5\times$ this value, $I_{\rm det}=5\times 10^{-4}\,{\rm m}\,{\rm
  s}^{-1},$ can be used to directly constrain the relevant deflector and include it in the ephemeris model.
When normalized by $v_c$ yields $I_{\rm  det}=10^{-8}$ at 10\% accuracy across the main belt.

@@

The first element adopted for $\vecq$ is the shift $\Delta a$, derivable from the orbital energy $E=-1/2a.$
\begin{eqnarray}
  \Delta E & = &  \vecv \cdot \vecI + O(I^2) \\
  \quad \Rightarrow \quad \Delta a & = & 2\left(v_r I_r + v_\phi I_\phi\right) = -2e_\perp I_r + 2(1+e_\parallel) I_\phi.
  \label{eq:da}
\end{eqnarray}
The initial angular momentum $\mathbf{L} = \sqrt{1-e^2}\,\zhat \approx \zhat$ is altered by
\begin{eqnarray}
  \Delta \mathbf{L} & = & \vecr \times \vecI = (1-e_\parallel) I_\phi \zhat - (1-e_\parallel) I_z \phat \\
  \quad \Rightarrow \quad \Delta \Lhat_\parallel & = & -2e_\perp I_z
  \label{eq:dLpar}\\
  \Delta \Lhat_\perp & = & -(1-e_\parallel)  I_z
                  \label{eq:dLperp}
\end{eqnarray}
We adopt the changes in direction $(\Delta\Lhat_x, \Delta\Lhat_y)$ as the next two elements of $\vecq$ specifying the inclination and ascending node $\Omega$ of the perturbed orbit.  These are a rotation of $(\Delta\Lhat_\parallel, \Delta\Lhat_\perp)$ by the angle $\tau.$  The conversion from the first line above to the subsequent two makes use of the decomposition $\phat=\uhat_\perp + 2e_\perp \uhat_\parallel$ derivable from the value of $\phi$ in Equations~(\ref{eq:kepler}).

@@

We now have all the information we need to create the $\matA$ matrix.
Rotating the $\vecq$ elements back into the Cartesian frame, we
obtain, to $O(e)$,
\begin{equation}
  \left( \begin{array}{c}
           \Delta a \\ \Delta\tau \\ \Delta e_x \\ \Delta e_y \\ \Delta \Lhat_x \\ \Delta\Lhat_y
         \end{array}\right) = \left( \begin{array}{ccc}
 2e\sin\tau & 2 + 2e\cos\tau & 0 \\
 -2+e\left[1.5\cos\tau-3(T-t_i)\sin\tau\right] & e\sin\tau -3(1+e\cos\tau)(T-t_i) & 0\\
 -\cos\tau-2e\sin^2\tau & 2\cos\tau - 1.5e\sin 2\tau & 0\\
  -\sin\tau + e\sin 2\tau & 2\sin\tau-3e\cos^2\tau & 0 \\
0 & 0 & \sin\tau + 0.5 e \sin 2\tau \\
 0 & 0 &  -\cos\tau +e(3-\cos 2\tau)/2 \\
 \end{array}\right)
  \left(\begin{array}{c} I_r \\ I_\phi \\ I_z \end{array}\right)
\end{equation}
We can now do the time integral and sum over impulse directions in \eqq{eq:Cqjk}, and average over $\tau$ to average over the orbital phase of the impulses. The diagonal elements are
\begin{eqnarray}
  \Var(a) & = & 4T \nIphi \nonumber\\
  \Var(\tau) & = & 4T\nIr + 3T^3\nIphi \nonumber\\
  \Var(e_x) =  \Var(e_y) & = & \nIr/2+2\nIphi \nonumber\\
  \Var(\Lhat_x) =  \Var(\Lhat_y) & = & \nIz/2
\label{eq:Cq}
\end{eqnarray}
All of the off-diagonal terms (covariances) are either zero or $O(e)\left\langle nI^2\right\rangle,$ and they result in observable consequences that are smaller by $\left\langle e^2 \right\rangle$ than the diagonal terms' contributions.

>>>>

One of our tasks will be to derive, from the known asteroid population, the rate (per tracer) of encounters vs the imparted impulse,
\begin{equation}
  \frac{dN}{dt\,dI_r\,dI_\phi\,dI_z},
  \label{eq:dN}
\end{equation}
where the components of $\vecI$ are given in the cylindrical basis
vectors about the asteroid's position at the impulse.
We will estimate this function from the known population,
approximating it as constant within each of three subsets of tracer MBAs:
the Inner, Middle, and Outer regions bounded by the 4:1, 3:1,
5:2, and 2:1 mean-motion resonances with Jupiter.
We will further assume that the impulses on a given target are drawn
independently from this distribution, \ie\ a Poisson process defined
by this rate.  In this case, the only properties of the impulse
distribution that we need are its second moments for components $\alpha \in \{r,\phi,z\}$
\begin{equation}
  \left \langle n I_\alpha^2 \right\rangle \equiv \int d^3I \frac{dN}{dt\,dI_r\,dI_\phi\,dI_z} I_\alpha^2.
\label{eq:nvsq}
\end{equation}
The average of the cross terms $I_rI_z$ and $I_\phi I_z$ vanishes if
the deflector distribution is symmetric in inclination as expected, and we find numerically that the mean $I_rI_\phi$ is small enough to ignore.

From this knowledge of the impulse distribution, our goal is to obtain
the covariance matrix of the deviations of the target's position from
the initial orbit, after some time $T.$ The position observables are
the range, plus the celestial latitude and longitude of the target.  We will simplify our results by assuming a heliocentric observer, so that the observational position vector is $\vecp\equiv (r,\phi,\theta=z/r).$ The quantity we seek is the covariance matrix $\covm^p$ of the observations attributable to the accumulated gravitational perturbations:
\begin{equation}
  \covm^p  \equiv  \left\langle  \Delta\vecp \Delta\vecp^T\right\rangle,  \label{eq:Cp}
\end{equation}
where the angle brackets indicate an average over possible
realizations of the impulse history, and over the mean anomaly $M$ at the time of observation.  To do so, we will introduce an intermediate set of 6 parameters $\vecq$ describing the deviations of the orbital elements induced by the impulses.  The $\vecq$ components will be selected to be equal to zero before any impulses are applied, and each responds linearly to $\vecI$ at $|I|\ll 1.$
We will derive the matrix $\matA$ that describes the orbital-element shifts at time $T$ that arise from an impulse at time $t_i$: 
\begin{equation}
  \Delta \vecq(e_0,T,t_i) = \matA(e_0, T, t_i) \vecI.
  \label{eq:A}
\end{equation}
In our first-order perturbation theory, $\vecq(T)$ will be the sum of the $\Delta\vecq_i$ imparted by all impulses applied at times $0<t_i<T.$  Because the impulses are uncorrelated, $\vecq$ will therefore be the result of a random walk.  The distribution will have a covariance matrix $\covm^q(T) \equiv \left\langle \vecq(T) \vecq^T(T) \right\rangle$ whose elements are
\begin{eqnarray}
  \covm^q_{jk}(e_o,T) & = & \left\langle \sum_{i,\gamma} \matA_{j\gamma}(e_0,T,t_i) \matA_{k\gamma}(e_0,T,t_i) I^2_{i,\gamma} \right\rangle \\
           & = & \sum_\gamma \int dt_i \matA_{j\gamma}(e_0,T,t_i) \matA_{k\gamma}(e_0,T,t_i) \left\langle n I_\gamma^2\right\rangle.
\label{eq:Cqjk}
\end{eqnarray}
In the first line, the sum $i$ runs over the impulses and $\gamma$ runs over the components $r,\phi,z$ of the impulse. The second line evaluates the expectation value of averaging over realizations of the random walk of impulses, exploiting the independence of the individual impulses from each other.

The last element of our calculation will be a conversion from the
element shifts $\vecq$ into the observed displacements $\Delta\vecp$
at time $T.$  In linear perturbation theory this will again be expressible as a matrix
\begin{eqnarray}
\label{eq:B}
  \Delta\vecp(e_0,T) & = & \matB(e_0,T) \vecq(T) \\
  \Rightarrow \quad 
  \covm^p_{\alpha\beta}(e_0,T) & = & \sum_{jk} B_{\alpha j}(e_0,T) B_{\beta k}(e_0,T) \covm^q_{jk}(T) 
\label{eq:covp}
\end{eqnarray}
which we would wish to average over the phase of the MBA orbit or mean anomaly $M$ at the time of observation $T.$

@@

\section{Impulse distribution}
\label{sec:impulse}
\subsection{Derivation}
Under the assumption that the distributions of masses and orbital
elements of MBAs are separable, the quantities needed to describe the
random walk of the asteroid's orbits can be expressed as an integral
over the absolute magnitude $H$ of the deflectors, the velocity $v$,
impact parameter $b,$ and polar/azimuthal angles $\theta,\phi$ of the
unit vector of the impulse direction $\bhat$:
\begin{eqnarray}
  \left\langle n I_\gamma^2 \right\rangle & = & \int
                                              dH\,dv\,db\,d\theta\,d\phi
                                             \frac{dN}{dH} \frac{dn}{dv\,db\,d\theta\,d\phi}
                                              I^2_\gamma(H,v,b,\theta\phi)
  \\
  & = & \int dH \frac{dN}{dH} \int dv
        \frac{dn}{dA\,dv\,d\theta\,d\phi} \hat b_\gamma^2(\theta,\phi)
        \int_{b_{\rm min}}^{b_{\rm max}} 2\pi b\,db\,
        \left(\frac{2\sigma_M(H)}{bv}\right)^2 \\
  & = & 8\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int dv
        \frac{dn}{dA\,dv} v^{-2} \left\langle\hat b_\gamma^2(v)\right\rangle
        \log(b_{\rm max}/b_{\rm min}).
\end{eqnarray}
The second row adopts the (numerically verified) assumption that the
distribution of impact parameter $b$ will always be $\propto dA=2\pi
b\,db$ in the range of $b$ of interest to us.  The mean rate of encounters
between a deflector-tracer pair, as a function of the impact velocity
and geometry, is $dn/dA\,dv\,d\theta\,d\phi,$ which we will determine
through numerical measurements with the orbits of all known MBAs. The
$H$ distribution of MBAs is $dN/dH.$ We denote as $\sigma_M(H)$
the RMS error in the mass of MBAs at $H$ from the value used in the ephemeris model.
We use $\sigma_M$ rather than $M$ since we are seeking the size of
deviations from the ephemeris model.  For MBAs that are not included
in the ephemeris model, $\sigma_M=M(H),$ the full mass of the deflector.

@@

If we crudely choose some threshold
$I_{\rm det}$ of impulse as being large enough to generate high-$S/N$
deflections on its tracer, then our condition becomes $2M(H)/bv <
I_{\rm det},$ or $b_{\rm min}=2M(H)/vI_{\rm det}.$  We then have
\begin{equation}
   \left\langle n I_\gamma^2 \right\rangle = 
2\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int_{\sqrt{2M(H)/I_{\rm
      det}}}^\infty \frac{dv}{v^2} 
        \frac{dn}{dA\,dv}  \left\langle\hat b_\gamma^2(v)\right\rangle
        \log \left[v^2 I_{\rm det}/2M(H)\right].
        \label{eq:nI2}
      \end{equation}
The lower bound on the $v$ integral marks the speed below which
$b_{\rm max}<b_{\rm min}.$  Since $b_{\rm min}$ and $b_{\rm max}$
appear logarithmically, the estimated diffusion is relatively insensitive to choices of
their values.

@@

From this full MBA list, we select a random subset of 5000 objects
from each of the Inner, Middle, and Outer belts to serve as a sample
of ``tracers.''  We record all of the passages of any tracer MBA
within 0.03~AU of any other ``deflector'' MBA (drawn from the full
catalog of 1.26~million) over a 10-year period.  The 5000 tracers per
belt region are enough to be representative of the statistics of
the region, but few enough that the calculation can be done easily
on a laptop computer.
We will estimate the constituents of  $\langle nI_\gamma^2\rangle$ in
\eqq{eq:nI2} separately for each of the nine combinations of (tracer region, deflector region).

@@

The leapfrog integrator advances all 1.26 million state vectors by 2
days, and the process is repeated.  We continue searching for pairs at
2-day intervals until we have recorded 10 years' worth of encounters---17.7 million
total impulses, or roughly 1000 encounters per tracer.

@@

For each pair of tracer-deflector regions (\eg\ Inner-Inner), we combine the list of
events with the counts of candidate tracers and deflectors to
calculate $dn/dA\,dv,$ the event rates per eligible pair; and the mean
geometry of the encounter, $\langle \hat b^2_\gamma \rangle,$ as a
function of $v$, all plotted in Figure~\ref{fig:rtz}.
Here it is apparent that the $\hat b_\phi$ component is
larger than $\hat b_z$ and $\hat b_r,$ but there is significant
variation with MBA class and with $v$.  An important result is that
the quantity $v^{-2} (dN/dA\,dv)$ is always well-behaved as $v\rightarrow 0,$
so we do not need any special treatment to sum the integrals in
\eqq{eq:nI2}.  The $v$ integrand peak is in the range $0.03<v/v_c<0.1.$

@@

To give some context, a
comprehensive investigation of MBA masses from mutual scattering
events by \citet{goffin} reports typical $\sigma_M\approx
5\times10^{-13}$ for 30 MBAs. But we can expect substantial improvement
in the LSST era when a millions of potential tracer asteroids attain
mas-level astrometry. \citet{negin} use a simulated LSST sequence of observations to estimate the impulse (and mass) uncertainties from analyzing LSST's data on mutual encounters of the known asteroids. The tracer MBAs for those events are essentially an ubiased selection from the known MBAs, and the left-hand plot of Figure~\ref{fig:sigmaI} shows the distribution of the uncertainty $\sigma_I$ derived from each tracer's LSST data after marginalizing over its initial state vector.  This histogram marginalizes over the apparent magnitude and observing conditions of the tracer, the date of the encounter, and the direction $\bhat$ of the impulse.  The right-hand plot of Figure~\ref{fig:sigmaI} forecasts the distribution of $\sigma_M$ that we can expect for deflectors.  This forecast is made as follows: we treat each of the 5000 tracer MBAs that we have selected from each region as deflectors, since we have a list of all of their encounters with known MBAs with $b<0.03$~AU.  For each such encounter, we select at random one $\sigma_I$ from the distribution found by \citet{negin} (Figure~\ref{fig:sigmaI}, left), and calculate $\sigma_M=(bv/2)\sigma_I.$  For each deflector, we then calculate (a) the individual encounter with lowest $\sigma_M$; (b) the combined constraint $\left(\sum \sigma_{M,i}^2\right)^{-1/2}$ from the 10 lowest $\sigma_M$ encounters for that deflector; and (c) the combined constraint for all $b<0.03$~AU encounters (typically $\sim1000$ of them).  The right-hand plot of Figure~\ref{fig:sigmaI} gives the histograms of these three measures of $\sigma_M.$  We see that most of the available information on any given deflector's mass comes from its 10 most-informative encounters.  It would be computationally very easy to solve for the masses of $O(10^4)$ MBAs by including the deflections they impart on $O(10^5)$ other MBAs in an ephemeris fit.
The RMS $\sigma_M$ using LSST data from the 10 most-informative tracer encounters with a given deflector is estimated as $\sigma_M=2.6\times10^{-14}M_\odot.$ There are $\approx450$ MBAs with mass above this forecasted RMS $\sigma_M.$  This is a conservative estimate in that we ignore any observational information before LSST, and we also do not consider the information increase that will come from tracking the $\approx10$-fold increase in number of known MBAs that LSST will enable.

\subsubsection{Individually detectable impulses}
Examining the $\sigma_I$ distribution forecasted for LSST data on tracers (Figure~\ref{fig:sigmaI}), we find that for impulses occurring near the midpoint of the survey, $\approx80\%$ of all cases yield $\sigma_I<10^{-4}\,{\rm m}\,{\rm
  s}^{-1},$ some an order of magnitude lower.  We conservatively
assume that any encounter producing  an impulse $>5\times$ this value, $I_{\rm det}=5\times 10^{-4}\,{\rm m}\,{\rm
  s}^{-1},$ can be used to directly constrain the relevant deflector and include it in the ephemeris model.
When normalized by $v_c,$ we have $I_{\rm  det}=10^{-8}$ at 10\% accuracy across the main belt.

@@

The first element adopted for $\vecq$ is the shift $\Delta a$, derivable from the orbital energy $E=-1/2a.$
\begin{eqnarray}
  \Delta E & = &  \vecv \cdot \vecI + O(I^2) \\
  \quad \Rightarrow \quad \Delta a & = & 2\left(v_r I_r + v_\phi I_\phi\right) = -2e_\perp I_r + 2(1+e_\parallel) I_\phi.
  \label{eq:da}
\end{eqnarray}
The initial angular momentum $\mathbf{L} = \sqrt{1-e^2}\,\zhat \approx \zhat$ is altered by
\begin{eqnarray}
  \Delta \mathbf{L} & = & \vecr \times \vecI = (1-e_\parallel) I_\phi \zhat - (1-e_\parallel) I_z \phat \\
  \quad \Rightarrow \quad \Delta \Lhat_\parallel & = & -2e_\perp I_z
  \label{eq:dLpar}\\
  \Delta \Lhat_\perp & = & -(1-e_\parallel)  I_z
                  \label{eq:dLperp}
\end{eqnarray}
We adopt the changes in direction $(\Delta\Lhat_x, \Delta\Lhat_y)$ as two elements of $\vecq$ specifying the inclination and ascending node $\Omega$ of the perturbed orbit.  These are a rotation of $(\Delta\Lhat_\parallel, \Delta\Lhat_\perp)$ by the angle $\tau.$  The conversion from the first line above to the subsequent two makes use of the decomposition $\phat=\uhat_\perp + 2e_\perp \uhat_\parallel$ derivable from the value of $\phi$ in Equations~(\ref{eq:kepler}).

@@

We now have all the information we need to create the $\matA$ matrix.
Rotating to $\vecq=\{\Delta a, \Delta \tau, \Delta e_x, \Delta e_y, \Delta\Lhat_x, \Delta\Lhat_y\}$ and with $\vecI=\{I_r, I_\phi, I_z\}$ we have, to $O(e)$,
\begin{equation}
         \matA(e,T,t_i) = \left( \begin{array}{ccc}
 2e\sin\tau & 2 + 2e\cos\tau & 0 \\
 -2+e\left[1.5\cos\tau-3(T-t_i)\sin\tau\right] & e\sin\tau -3(1+e\cos\tau)(T-t_i) & 0\\
 -\cos\tau-2e\sin^2\tau & 2\cos\tau - 1.5e\sin 2\tau & 0\\
  -\sin\tau + e\sin 2\tau & 2\sin\tau-3e\cos^2\tau & 0 \\
0 & 0 & \sin\tau + 0.5 e \sin 2\tau \\
 0 & 0 &  -\cos\tau +e(3-\cos 2\tau)/2 \\
 \end{array}\right)
\end{equation}
We can now do the time integral and sum over impulse directions in \eqq{eq:Cqjk}, and average over $\tau$ to average over the orbital phase of the impulses. The diagonal elements of $\covm^q$ are
\begin{eqnarray}
  \Var(a) & = & 4T \nIphi \nonumber\\
  \Var(\tau) & = & 4T\nIr + 3T^3\nIphi \nonumber\\
  \Var(e_x) =  \Var(e_y) & = & \nIr/2+2\nIphi \nonumber\\
  \Var(\Lhat_x) =  \Var(\Lhat_y) & = & \nIz/2
\label{eq:Cq}
\end{eqnarray}
All of the off-diagonal terms (covariances) are either zero or $O\left(e\left\langle nI^2\right\rangle\right),$ and they result in observable consequences that are smaller by $\left\langle e^2 \right\rangle$ than the diagonal terms' contributions.

% \begin{equation}
%   \matC^2 = \left(\begin{array}{cccccc}
%     4T \nIphi & -3T^2\nIphi & e\left(\nIr + 5\nIphi\right) & e\nIr & 0 & 0 \\
% -3T^2\nIphi & 4\nIr + 3T^3\nIphi & -e(11T\nIr/4 + 15T^2\nIphi/4) & e\left[(2T-3T^2/4)\nIr+(-T+9T^2/4)\nIphi\right] & 0 & 0\\
%  e\left(\nIr + 5\nIphi\right) & -e(11T\nIr/4 + 15T^2\nIphi/4) & \nIr/2+2\nIphi & 0 & 0 & 0 \\
%  e\nIr &  e\left[(2T-3T^2/4)\nIr+(-T+9T^2/4)\nIphi\right]  & 0 & \nIr/2+2\nIphi & 0 & 0 \\
% 0 & 0 & 0 & 0 & \nIz/2 & 0 \\
% 0 & 0 & 0 & 0 & 0 \nIz/2 \\
%\end{array}\right)
% \end{equation}
 \subsection{Astrometric and ranging shifts from elements shifts}
\label{sec:observe}
The radial coordinate of a Keplerian orbit is $r=a(1-e\cos E),$ and an expansion of the eccentric anomaly $E$ in powers of $e$ is
\begin{equation}
  \label{eq:Ee2}
  \cos E  =  e\cos M - (e \sin M)^2 + O(e^3).
\end{equation}
From this we can derive the perturbations to $r$ as a function of our chosen basis for perturbations to the orbital elements.  Making use of Equations~(\ref{eq:ecosM}) and (\ref{eq:esinM}) we obtain
\begin{equation}
  \Delta r = (1-e\cos\tau) \Delta a + e\sin\tau \Delta\tau +  (-\cos\tau -2e\sin^2\tau) \Delta e_x + (-\sin\tau+e\sin 2\tau) \Delta e_y.
\end{equation}
If we square this expression, average over the orbital phase $\tau$ of the observation, and retain leading order in $e$, we get the variance of the radial coordinate from the unperturbed orbit.  All of the terms involving covariances between orbital elements either average to zero over an orbit, or are suppressed by $e^2$ relative to other terms.  We are left with
\begin{eqnarray}
  \Var(r) & \approx & \Var(a) + \frac{1}{2}\Var(e_x) + \frac{1}{2}\Var(e_y) + \frac{\langle e^2\rangle}{2} \Var(\tau) \\
          & \approx & \frac{T}{2} \nIr + \left( 6T + \frac{3\langle e^2 \rangle}{2} T^3\right) \nIphi.
                      \label{eq:varr}
\end{eqnarray}
The second line substitutes the results from Equations~(\ref{eq:Cq}).  We retain the final term because, as $T>>1,$ this precession term that scales with $\langle e^2 \rangle T^3$ becomes comparable to or larger than the other terms that scale as $T$.  
The heliocentric azimuthal angle $\phi$ is given by \eqq{eq:phi}.  We can similarly square the perturbation and average over orbital phase and population factors to get
\begin{eqnarray}
  \Var(\phi) & \approx & \Var(\tau) + 2\Var(e_x) + 2\Var(e_y) \\
             & \approx & 6T\nIr + \left(8T+3T^3\right) \nIphi.
                         \label{eq:varphi}
\end{eqnarray}
The heliocentric latitude (or equivalently, polar angle) has a deviation
\begin{eqnarray}
  \Delta\theta & = & -\Delta\Lhat_x \frac{x}{r} - \Delta\Lhat_y \frac{y}{r} \\
  \Rightarrow \quad \Var(\theta) & \approx & \langle\cos^2\tau\rangle \Var(\Lhat_x) + \langle\sin^2\tau\rangle \Var(\Lhat_y) \\
               & \approx & \frac{T}{2} \nIz.
                           \label{eq:vartheta}
\end{eqnarray}
